The Weakly Nonlinear Magnetorotational Instability in a local geometry

### 2017-03-12 paper/thingap_ref_response/response_to_referee.tex

\begin{enumerate}				
\item One of the main ideas behind the “weakly nonlinear” approach is that near marginal stability there is only one or a very small band of unstable modes that self interact. There is a substantial body of work in the fusion plasma community that considers coupling to damped modes as an important ingredient in saturation. The names to look for are Hatch, Makwana, and Terry. I am not suggesting the authors redo their analysis, to include damped modes, but they should comment on this alternative, perhaps in the Introduction.

\textbf{Coupling to damped modes as explored in the plasma literature is not analytically possible in the walled TC flow, but can be analyzed in the shearing box when ambipolar diffusion is considered, as described in the new section of the text.}

\item With regard to saturation through modification of the base flow, in the TC problem studied in these papers, the rotation law is enforced through viscous coupling of the interior fluid to the bounding cylinders. This introduces a timescale, the viscous diffusion time, with which the momentum transport driven by perturbations must compete. In an accretion disk the rotation law is enforced at every point directly by gravity. The authors should comment on how important they think this difference is to saturating the mode.

\textbf{In Paper II, the rotation law is enforced through viscous coupling to the boundaries. In Paper I, however, the rotation law is imposed directly at every point by the shear. We have noted this point in the text. See also the newly extended discussion of the local flow with shear periodic boundary conditions.}

\item The influence of boundary conditions and boundary layers is strong. To call out some examples: In Paper I, the amplitude $\alpha$ converges to the constant value $(b/c)^{1/2}$, independent of z. Is this really relevant to a lab experiment with vertical endcaps, or to a stratified disk? Have the authors experimented with a finite cylinder, even a very tall one? Likewise, the radial boundaries seem to play a role that a softer physical boundary might not, for example the boundary layers in Figure 2 of Paper II (in this case it would be useful to show blowups of the solution, as in Figure 4, and present evidence for numerical convergence). I would like to see more discussion of the role of these boundary conditions versus physical boundary conditions. Again, it is OK to model a lab experiment – but the connection to a natural system should be kept in mind.

@@

\textbf{The neutral mode is a spatially constant azimuthal magnetic field. It cannot nonlinearly interact with any modes -- damped or otherwise -- because all of the nonlinearities involve derivatives. This remains true when ambipolar diffusion is considered. We have added two sentences to Section 3 to clarify this point.}

%Neutral mode doesn’t exist in cylindrical geometry, yet we find same GLE -- can we just argue that it’s not important for saturation? 
%The beta mode arises because a constant By solves the equations. However, this cannot non-linearly couple to any other modes, because all NL terms in the problem involve derivatives. Thus, there is no coupling to it, even though we discard couplings between the other damped eigenmode families in the thin-gap MHD system (that is, the non-MRI modes)

\item The saturated Bz profile shown in Figure 5 of Paper I is intriguing in light of the fact that the authors achieved instability by dialing down Bz from the marginally stable value. Would a radial average of the new Bz, which is stronger at the edges and weaker in the middle, reveal that it is marginally stable? Do the results bear on whether the MRI is a dynamo? There is no counterpart figure in Paper II; the authors should comment on dynamo action here, too.

\textbf{Yes, the saturated Bz is marginally stable, and we have added a sentence to that effect in the text. We cannot comment on dynamo action in either paper because of the symmetries in our setup. No dynamo can arise from an axisymmetric system (e.g. Cowling's anti-dynamo theorem).}

### 2017-03-12 paper/thingap_ref_response/thingap_mri_paper_ref_response_apj.tex
many more additions to text

\subsection{\edit1{Shearing box and ambipolar diffusion}}

Many studies of the MRI consider the instability in a shearing box, i.e. a wall-less local approximation that is meant to represent a small section of a disk. The shearing box obtains when Equations \ref{eqset1} - \ref{eqset4} are subjected to shear periodic radial boundary conditions rather than Equation \ref{eq:wallbcs}. The periodic nature of the shearing box allows us to decompose the fluid perturbations into Fourier modes proportional to $e^{i k_x x + i k_z z}$. This makes the shearing box MRI straightforward to treat analytically. However, as noted above, the fastest-growing linear MRI modes in the shearing box are also exact solutions of the nonlinear MRI equations -- that is, $J(\hat{\psi_0}, \hat{\psi_0}) = J(\hat{\psi_0}, \nabla^2 \hat{\psi_0}) = 0$ for $\hat{\psi_0} \propto e^{i k_x x + i k_z z}$. While this may be an appealing trait for analytic simplicity, it leads to the unphysical conclusion that the fastest growing modes will never nonlinearly interact \citep{Goodman:1994ul}. This `nonlinear property' will not be satisfied for two MRI modes with nonparallel wavenumbers, but with vertically periodic boundary conditions and a vertical background magnetic field the most unstable mode has a strictly axial vertical wavenumber. Thus a formal weakly nonlinear analysis cannot be conducted, as the most unstable mode will never nonlinearly interact with itself or its complex conjugate. Similarly, we cannot analytically examine interactions between MRI channel modes and damped eigenmodes belonging to other wave families. This is analytically tracked for other plasma instabilities by tracking the amplitudes of growing, marginal, and damped eigenmodes simultaneously \citep[e.g.][]{Makwana:2011}. While the shearing box approximation allows the projection of the perturbed MRI equations into the basis set of linear eigenmodes, nonlinear coupling between modes will remain zero.

The nonlinear property of primary MRI modes in the shearing box motivates the addition of radial boundaries, such that the nonlinear evolution of the weakly nonlinear MRI can be properly considered. It also begs the question of whether some additional nonlinear mechanism can be introduced such that the fastest-growing modes are no longer nonlinear solutions to the shearing box equations. It has already been shown that the Hall effect does not negate the nonlinear property of primary MRI modes \citep{Kunz:2013}. However, it seems to have been overlooked in the literature that these linear modes are \textit{not} solutions of the nonlinear ambipolar diffusion term, which is proportional to

@@

Furthermore, the radial wavenumber of the fastest-growing linear MRI mode in a shearing box with ambipolar diffusion is nonzero when a constant azimuthal background field is considered in addition to an axial one \citep{Kunz:2004ib}. This means that, in the presence of ambipolar diffusion, we can derive the weakly nonlinear envelope equation for the MRI in the shearing box. Ambipolar diffusion adds both linear and nonlinear terms to Equations \ref{eq:ordere} to \ref{eq:ordere3}, but does not change their $Z$ or $T$ dependence. The constant azimuthal background field component does not contribute to any other terms in the local MRI equations. Thus, the slow-scale evolution of the MRI in a shearing box with ambipolar diffusion is also governed by a Ginzburg-Landau equation.

The Ginzburg-Landau form of the amplitude equation can be found in any system with Euclidean symmetry and a quadratic maximum in growth rate with respect to the wavenumber \citep{Hoyle:2006}. In this case, the Euclidean symmetry comes from axisymmetry in the $x$-$z$ plane, and the quadratic maximum is a consequence of the linear dispersion relation given in Appendix~\ref{app:dispersion}. In Paper II, we show that the same symmetry occurs in the axisymmetric global geometry as well.
The Ginzburg-Landau equation arises due to symmetries in the local MRI equations, irrespective of the boundary conditions to which they are subjected. This means that the local MRI is able to saturate via nonlinear mode interaction so long as the primary MRI modes are not exact solutions of the nonlinear terms. This can be achieved by considering the effects of ambipolar diffusion when the boundary conditions are shear periodic, or by enforcing wall-like radial boundary conditions. Both constructions require the most unstable mode to have nonconstant radial structure. Physically, this radial variation impedes the free exchange of angular momentum facilitated by the uniform stretching of channel modes.  %So long as our system permits the most unstable mode to nonlinearly interact with

@@

>>>>

\subsection{\edit1{Shearing box and ambipolar diffusion}}\label{sec:SBAD}

\edit1{Many studies of the MRI consider the instability in a shearing box, i.e. a wall-less local approximation that is meant to represent a small section of a disk. The shearing box obtains when Equations \ref{eqset1} - \ref{eqset4} are subjected to shear periodic radial boundary conditions rather than Equation \ref{eq:wallbcs}. The periodic nature of the shearing box allows us to decompose the fluid perturbations into Fourier modes proportional to $e^{i k_x x + i k_z z}$. This makes the shearing box MRI straightforward to treat analytically. However, as noted above, the fastest-growing linear MRI modes in the shearing box are also exact solutions of the nonlinear MRI equations -- that is, $J(\hat{\psi_0}, \hat{\psi_0}) = J(\hat{\psi_0}, \nabla^2 \hat{\psi_0}) = 0$ for $\hat{\psi_0} \propto e^{i k_x x + i k_z z}$. While this may be an appealing trait for analytic simplicity, it leads to the unphysical conclusion that the fastest growing modes will never nonlinearly interact \citep{Goodman:1994ul}. This `nonlinear property' will not be satisfied for two MRI modes with nonparallel wavenumbers, but with vertically periodic boundary conditions and a vertical background magnetic field the most unstable mode has a strictly axial vertical wavenumber. Thus a formal weakly nonlinear analysis cannot be conducted, as the most unstable mode will never nonlinearly interact with itself or its complex conjugate. Similarly, we cannot analytically examine interactions between MRI channel modes and damped eigenmodes belonging to other wave families. This is analytically tracked for other plasma instabilities by tracking the amplitudes of growing, marginal, and damped eigenmodes simultaneously \citep[e.g.][]{Makwana:2011}. While the shearing box approximation allows the projection of the perturbed MRI equations into the basis set of linear eigenmodes, nonlinear coupling between modes will remain zero.}

\edit1{The nonlinear property of primary MRI modes in the shearing box motivates the addition of radial boundaries, such that the nonlinear evolution of the weakly nonlinear MRI can be properly considered. It also begs the question of whether some additional nonlinear mechanism can be introduced such that the fastest-growing modes are no longer nonlinear solutions to the shearing box equations. It has already been shown that the Hall effect does not negate the nonlinear property of primary MRI modes \citep{Kunz:2013}. However, it seems to have been overlooked in the literature that these linear modes are \textit{not} solutions of the nonlinear ambipolar diffusion term, which is proportional to}

@@

\edit1{Furthermore, the radial wavenumber of the fastest-growing linear MRI mode in a shearing box with ambipolar diffusion is nonzero when a constant azimuthal background field is considered in addition to an axial one \citep{Kunz:2004ib}. This means that, in the presence of ambipolar diffusion, we can derive the weakly nonlinear envelope equation for the MRI in the shearing box. Ambipolar diffusion adds both linear and nonlinear terms to Equations \ref{eq:ordere} to \ref{eq:ordere3}, but does not change their $Z$ or $T$ dependence. The constant azimuthal background field component does not contribute to any other terms in the local MRI equations. Thus, the slow-scale evolution of the MRI in a shearing box with ambipolar diffusion is also governed by a Ginzburg-Landau equation.}

\edit1{The Ginzburg-Landau form of the amplitude equation can be found in any system with Euclidean symmetry and a quadratic maximum in growth rate with respect to the wavenumber \citep{Hoyle:2006}. In this case, the Euclidean symmetry comes from axisymmetry in the $x$-$z$ plane, and the quadratic maximum is a consequence of the linear dispersion relation given in Appendix~\ref{app:dispersion}. In Paper II, we show that the same symmetry occurs in the axisymmetric global geometry as well.
The Ginzburg-Landau equation arises due to symmetries in the local MRI equations, irrespective of the boundary conditions to which they are subjected. This means that the local MRI is able to saturate via nonlinear mode interaction so long as the primary MRI modes are not exact solutions of the nonlinear terms. This can be achieved by considering the effects of ambipolar diffusion when the boundary conditions are shear periodic, or by enforcing wall-like radial boundary conditions. Both constructions require the most unstable mode to have nonconstant radial structure. Physically, this radial variation impedes the free exchange of angular momentum facilitated by the uniform stretching of channel modes.}  %So long as our system permits the most unstable mode to nonlinearly interact with

@@

\edit1{The weakly nonlinear theory predicts that the amplitude of the system is bounded by the saturation amplitude $\alpha_s = \sqrt{b/c}$, where $b$ and $c$ are coefficients corresponding to the linear growth term and nonlinear term of the Ginzburg-Landau equation, respectively. The coefficient $b$ comes from the interaction between the background magnetic field and the linear MRI solution. The coefficient $c$ describes the third-order nonlinear interaction between terms in the perturbation series. Physically, we see that the saturation amplitude is controlled by the strength of the mode interaction within our finite band of unstable modes. We stress that while the third-order nonlinear terms in the walled TC flow are strongly influenced by the boundary layers, this is not generically true of the MRI system. Indeed, in the shearing box MRI with ambipolar diffusion (the case sketched out in Section \ref{sec:SBAD}), boundary layers impossible in the shear periodic flow. In this case the third-order nonlinear behavior of the system includes three-mode interactions from the cubic nonlinearity in the ambipolar diffusion term.}
